---
title: 'Assignment 9: Web APIs'
author: "Bharani Nittala"
date: "`r Sys.Date()`"
output:
  html_document:
    includes: null
  pdf_document: default
editor_options:
  chunk_output_type: console
---

### Overview

The New York Times web site provides a rich set of APIs, as described here: https://developer.nytimes.com/apis. I selected 'Movie Reviews' API for this purpose.   


### Load required libraries

Step 1 is to install and load required libraries to extract data from API

```{r libraries, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(eval = TRUE, results = FALSE)
library(tidyverse)
library(dplyr)
library(httr)
library(jsonlite)
library(kableExtra)
library(glue)
library(rmarkdown)

```

### Connect to the API

Raw data looks like this:
```{r , message=FALSE,echo= FALSE}
apikey <- 'JCzbi47kjdS5QtTfarLOCJEkeb6g5is0'
```



```{r , message=FALSE,results="asis"}
#Interface to read the JSON data
url <- paste("https://api.nytimes.com/svc/movies/v2/reviews/search.json?query=&api-key=",apikey, sep='')

#Convert JSON to r dataframe
raw_data<- fromJSON(url, flatten = TRUE) %>% data.frame()
```


### Read Data

```{r read, message=FALSE,results="asis"}
#raw_data %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"), fixed_thead = T)
paged_table(raw_data, options = list(rows.print = 5))

```

Let's fetch title and review from this data and analyze it. 

```{r , message=FALSE,results="asis"}
data_final <- raw_data %>% select(results.display_title,results.headline,results.critics_pick,results.byline)
new_names <- c('Title','Review','Critics_pick','Reviewer')
colnames(data_final) <- new_names
```


### Analysis

We now have a list of movies that are critically acclaimed that we can add to our list
```{r , message=FALSE,results="asis"}

#In here, let's extract only the review from the text  
data_final <- data_final  %>% separate(Review,c("Title_again", "Review"),"Review:")
data_final <- data_final %>% select(Title,Review,Critics_pick,Reviewer)
data_final %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"), fixed_thead = T)

```


### Conclusion

Ever wondered though if there could a reviewers bias? For example, my wife likes every movie she watches whereas it's hard to impress me unless there's depth in the story. Could there be bias among the reviewers as well ?

```{r compare, message=FALSE,results="asis"}

Critics <- data_final %>% group_by(Reviewer) %>% summarise(All_movies = n(),Acclaimed = sum(Critics_pick))
Critics_review <- Critics %>% filter(All_movies>1) %>% group_by(Reviewer) %>% summarise(acclaim_rate = round(Acclaimed/All_movies *100,0))

Critics_review %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed"), fixed_thead = T)
```

#### Seems like Ben is like me who is hard to impress whereas Glenn is more easy to impress. It's purely an hypothesis, for detailed analysis the confounding variables like movie genre, movie cast and other variables are required. Also, definitely more than 20 sample data points!
